---
title: "Milestone 4 - Denominators (Tidy and Join)"
format: html
date: "`r Sys.Date()`"
embed-resources: true
---

```{r}
library(tidyverse)
library(quarto)
library(arrow)
library(waldo)

source('functions/fxn_add_event_counts.R')

solution<-read_rds(here::here('milestones_dairy/data_milestones/solutions_week_05_pivotlonger.rds'))

```

## Denominators

Now that you can tidy data, pivot longer or wider, and then join tables to others, you are both unstoppable . . . and dangerous.

In this lesson we will explore one method of using very granular denominators in your reports.

We begin by reading in files for events, and denominators

```{r initial denominator generation, message==FALSE}

denominator_base<-read_parquet(here::here('data/intermediate_files/denominator_by_calendar_time_period.parquet'))
events_formatted <- read_parquet(here::here('data/intermediate_files/events_formatted.parquet'))

```


The first step is to filter your denominator according to the `calendar_time_period_type` that you want. 

The options for calendar periods are: **`r sort(unique(denominator_base$calendar_time_period_type))`**

The second step is to choose your lactation grouping. 
To simplify this first example will select a grouping without day of phase. 
(see the DOP section of How to Use denominaotrs report for day of phase examples)

The options for groups are: **`r sort(unique(denominator_base$deno_type))`**

Your final code should look something like this:

```{r, echo = TRUE}

denominator<-read_parquet(here::here('data/intermediate_files/denominator_by_calendar_time_period.parquet'))%>%
  filter(calendar_time_period_type %in% 'month')%>% #choose denominator period type
  filter(deno_type %in% 'lact_3+')%>% #choose the type of lactation group
  distinct()%>%
  select(-contains('dop'), -day_of_phase_group) # day of phase columns are not needed when dop is not part of grouping

```

```{r}
head(denominator)
```
### Step 2 - Summarize your numerator correctly

Follow the comments documenting the steps for nummerator summary. 

Make sure you understand them.  This is a functional example of what you will be asked to do in step 5 for an event or disease of your choice.


```{r, echo = TRUE}

mast_event_summary<-read_parquet(here::here('data/intermediate_files/events_formatted.parquet'))%>%
  
  #select event of interest
  filter(event %in% 'MAST')%>% #
  
  #add the event counts the way you like
  fxn_add_event_counts()%>% 
  mutate(event_count_lact = case_when(
    event_count_lact <4~paste0('Mast ', event_count_lact), 
    is.na(event_count_lact) ~ 'Unknown',
    TRUE~'Mast 4+'
  ))%>%
  
  #match the dates to the denominator
  mutate(date_time_period_start = floor_date(date_event, unit = 'month'))%>% 
  
  #use same  lactation grouping as denominator (0, 1, 2, 3+), 
  #and set location to match denominator (locations will depend of farm data structure)
  mutate(`Lactation Group` = lact_group, 
         location_lact_list = location_event) %>% 
  
  #group by all variables in the denominator, plus any extras that you want (like event count), and summarize
  group_by(location_lact_list, `Lactation Group`, 
            date_time_period_start, event_count_lact)%>% 
  summarize(ct_mast_events = n_distinct(id_animal))%>%
  ungroup()
  

```

### Step 3 - Join denominator to numerator

Be careful with these joins.  How do you know it joined correctly?
Add some code to this chunk to demonstrate that it performed the way you expected.

```{r, echo = TRUE}
final_data<-denominator%>%
  left_join(mast_event_summary)%>%
  mutate(ct_mast_events = replace_na(ct_mast_events, 0))%>%
  mutate(mast_rate = ct_mast_events/ct_animal_time_periods)

head(final_data)
```


### Step 4 - Look at results

```{r, echo = TRUE}

ggplot(final_data%>%filter(!(is.na(event_count_lact))))+
  geom_point(aes(x = date_time_period_start, y= mast_rate, color = `Lactation Group`))+
    geom_smooth(aes(x = date_time_period_start, y= mast_rate, 
                    color = `Lactation Group`, group =  `Lactation Group`, 
                    fill = `Lactation Group`),alpha = .2 , span = .6)+

  facet_grid(event_count_lact~location_lact_list, scales = 'free')



```
## Step 5 - your turn

Repeat the steps above for a disease of your choice. Then plot the trend of that disease incidence over time.

Be careful with your joins. Do not unintentionally create or delete data.

```{r}

#Pick Calendar Time Period Type and Lactation Group for denominator----

denominator2<-read_parquet(here::here('data/intermediate_files/denominator_by_calendar_time_period.parquet'))%>%
  filter(calendar_time_period_type %in% 'season')%>% #choose denominator period type
  filter(deno_type %in% 'lact_3+')%>% #choose the type of lactation group
  distinct()%>%
  select(-contains('dop'), -day_of_phase_group) # day of phase columns are not needed when dop is not part of grouping

#check data
head(denominator2)

mast_event_summary2<-read_parquet(here::here('data/intermediate_files/events_formatted.parquet'))%>%
  
  #select event of interest
  filter(event %in% 'MAST')%>% #
  
  #add the event counts the way you like
  fxn_add_event_counts()%>% 
  mutate(event_count_lact = case_when(
    event_count_lact <4~paste0('Mast ', event_count_lact), 
    is.na(event_count_lact) ~ 'Unknown',
    TRUE~'Mast 4+'
  )
  ) %>%
  
  #match the dates to the denominator (use lubridate season here, only made case_when season variable as practice)
  mutate(date_time_period_start= floor_date(date_event, unit = 'season'))%>% 
    mutate(date_time_period_start_season=case_when(
      month(date_time_period_start) %in% c(12, 1, 2) ~ "Winter",
  month(date_time_period_start) %in% c(3, 4, 5)  ~ "Spring",
  month(date_time_period_start) %in% c(6, 7, 8)  ~ "Summer",
  month(date_time_period_start) %in% c(9, 10, 11) ~ "Fall",
  TRUE ~ NA_character_
    )
    ) %>%

    #another way to make seasons with lubridate automatically (month to 12, 3, 6, or 9 and retains year) vs case_when (only downside of lubridate is you can't make your own seasonality groupings based on specific climate criteria as seasons)
       #mutate(date_time_period_start= floor_date(date_event, unit = 'season'))%>%

  
  #use same  lactation grouping as denominator (0, 1, 2, 3+), 
  #and set location to match denominator (locations will depend of farm data structure)
  mutate(`Lactation Group` = lact_group, 
         location_lact_list = location_event) %>% 
  
  #group by all variables in the denominator, plus any extras that you want (like event count), and summarize
  group_by(location_lact_list, `Lactation Group`, 
            date_time_period_start, event_count_lact)%>% 
  summarize(ct_mast_events2 = n_distinct(id_animal))%>%
  ungroup()

#join numerator and denominator data to make event case rates
final_data2<-denominator2%>%
  left_join(mast_event_summary2)%>%
  mutate(ct_mast_events2 = replace_na(ct_mast_events2, 0))%>%
  mutate(mast_rate = ct_mast_events2/ct_animal_time_periods)

head(final_data2) #check if it looks right

#graph event case rate
ggplot(final_data2%>%filter(!(is.na(event_count_lact))))+
  geom_point(aes(x = date_time_period_start, y= mast_rate, color = `Lactation Group`))+
    geom_smooth(aes(x = date_time_period_start, y= mast_rate, 
                    color = `Lactation Group`, group =  `Lactation Group`, 
                    fill = `Lactation Group`),alpha = .2 , span = .6)+

  facet_grid(event_count_lact~location_lact_list, scales = 'free')


####Graphing Practice
ggplot(final_data2%>%filter(!(is.na(event_count_lact))))+
  geom_point(aes(x = date_time_period_start, y= mast_rate, color = `Lactation Group`))+
    geom_smooth(aes(x = date_time_period_start, y= mast_rate, 
                    color = `Lactation Group`, group =  `Lactation Group`, 
                    fill = `Lactation Group`),alpha = .2 , span = .6)+

  facet_grid(event_count_lact~location_lact_list, scales = 'free')

```






## Part 2 - Extension

Think about your disease or event of interest. Standard denominators like the one above are often not good enough.

Use your new data wrangling skills to create the perfect denominator for your question.


NOTE: Denominators can be challenging. If you get stuck:

-   First articulate what you want to do, step by step, in English

-   Then add a blank code chunk for each step

-   Begin filling in code

-   Ask the group for help if you get completely stuck.

### Lameness Rates
```{r}

#Pick Calendar Time Period Type and Lactation Group for denominator----

denominator3<-read_parquet(here::here('data/intermediate_files/denominator_by_calendar_time_period.parquet'))%>%
  filter(calendar_time_period_type %in% 'month')%>% #choose denominator period type
  filter(deno_type %in% 'lact_3+')%>% #choose the type of lactation group
  distinct()%>%
  select(-contains('dop'), -day_of_phase_group) # day of phase columns are not needed when dop is not part of grouping


#check data
head(denominator3)

lame_event_summary<-read_parquet(here::here('data/intermediate_files/events_formatted.parquet'))%>%
  
  #select event of interest
  filter(event %in% 'LAME')%>% 
  #adding in filter for LAME to remove all TRIM ONLY Protocols would be great, but I believe the parquet files do not have events_formatted for me to filter out that data
  filter(protocol!='Trim Only') |>
  
  #add the event counts the way you like
  fxn_add_event_counts()%>% 
  mutate(event_count_lact = case_when(
    event_count_lact <4~paste0('Lame ', event_count_lact), 
    is.na(event_count_lact) ~ 'Unknown',
    TRUE~'Lame 4+'
  )
  ) %>%
  
  #match the dates to the denominator (use lubridate season here, only made case_when season variable as practice)
  mutate(date_time_period_start= floor_date(date_event, unit = 'month'))%>% 
    mutate(date_time_period_start_season=case_when(
      month(date_time_period_start) %in% c(12, 1, 2) ~ "Winter",
  month(date_time_period_start) %in% c(3, 4, 5)  ~ "Spring",
  month(date_time_period_start) %in% c(6, 7, 8)  ~ "Summer",
  month(date_time_period_start) %in% c(9, 10, 11) ~ "Fall",
  TRUE ~ NA_character_
    )
    ) %>%

    #another way to make seasons with lubridate automatically (month to 12, 3, 6, or 9 and retains year) vs case_when (only downside of lubridate is you can't make your own seasonality groupings based on specific climate criteria as seasons)
       #mutate(date_time_period_start= floor_date(date_event, unit = 'season'))%>%

  
  #use same  lactation grouping as denominator (0, 1, 2, 3+), 
  #and set location to match denominator (locations will depend of farm data structure)
  mutate(`Lactation Group` = lact_group, 
         location_lact_list = location_event) %>% 
  
  #group by all variables in the denominator, plus any extras that you want (like event count), and summarize
  group_by(location_lact_list, `Lactation Group`, 
            date_time_period_start, event_count_lact)%>% 
  summarize(ct_lame_events = n_distinct(id_animal))%>%
  ungroup()

#join numerator and denominator data to make event case rates
final_data3<-denominator3%>%
  left_join(lame_event_summary)%>%
  mutate(ct_lame_events = replace_na(ct_lame_events, 0))%>%
  mutate(lame_rate = ct_lame_events/ct_animal_time_periods)

head(final_data3) #check if it looks right

#graph event case rate
ggplot(final_data3%>%filter(!(is.na(event_count_lact))))+
  geom_point(aes(x = date_time_period_start, y= lame_rate, color = `Lactation Group`))+
    geom_smooth(aes(x = date_time_period_start, y= lame_rate, 
                    color = `Lactation Group`, group =  `Lactation Group`, 
                    fill = `Lactation Group`),alpha = .2 , span = .6)+

  facet_grid(event_count_lact~location_lact_list, scales = 'free')


####Graphing Practice
ggplot(final_data3%>%filter(!(is.na(event_count_lact))))+
  geom_point(aes(x = date_time_period_start, y= lame_rate, color = `Lactation Group`))+
    geom_smooth(aes(x = date_time_period_start, y= lame_rate, 
                    color = `Lactation Group`, group =  `Lactation Group`, 
                    fill = `Lactation Group`),alpha = .2 , span = .6)+

  facet_grid(event_count_lact~location_lact_list, scales = 'free')

```
