---
title: "Milestone 4 - Denominators (Tidy and Join)"
format: html
date: "`r Sys.Date()`"
embed-resources: true
---

```{r}
library(tidyverse)
library(quarto)
library(arrow)
library(waldo)

source('functions/fxn_add_event_counts.R')

solution<-read_rds(here::here('milestones_dairy/data_milestones/solutions_week_05_pivotlonger.rds'))

```

## Denominators

Now that you can tidy data, pivot longer or wider, and then join tables to others, you are both unstoppable . . . and dangerous.

In this lesson we will explore one method of using very granular denominators in your reports.

We begin by reading in files for events, and denominators

```{r initial denominator generation, message==FALSE}

denominator_base<-read_parquet(here::here('data/intermediate_files/denominator_by_calendar_time_period.parquet'))
events_formatted <- read_parquet(here::here('data/intermediate_files/events_formatted.parquet'))

```


The first step is to filter your denominator according to the `calendar_time_period_type` that you want. 

The options for calendar periods are: **`r sort(unique(denominator_base$calendar_time_period_type))`**

The second step is to choose your lactation grouping. 
To simplify this first example will select a grouping without day of phase. 
(see the DOP section of How to Use denominators report for day of phase examples)

The options for groups are: **`r sort(unique(denominator_base$deno_type))`**

Your final code should look something like this:

```{r, echo = TRUE}

denominator<-read_parquet(here::here('data/intermediate_files/denominator_by_calendar_time_period.parquet'))%>%
  filter(calendar_time_period_type %in% 'month')%>% #choose denominator period type
  filter(deno_type %in% 'lact_3+')%>% #choose the type of lactation group
  distinct()%>%
  select(-contains('dop'), -day_of_phase_group) # day of phase columns are not needed when dop is not part of grouping

```

```{r}
head(denominator)
```
### Step 2 - Summarize your numerator correctly

Follow the comments documenting the steps for numerator summary. 

Make sure you understand them.  This is a functional example of what you will be asked to do in step 5 for an event or disease of your choice.


```{r, echo = TRUE}

mast_event_summary<-read_parquet(here::here('data/intermediate_files/events_formatted.parquet'))%>%
  
  #select event of interest
  filter(event %in% 'MAST')%>% #
  
  #add the event counts the way you like
  fxn_add_event_counts()%>% 
  mutate(event_count_lact = case_when(
    event_count_lact <4~paste0('Mast ', event_count_lact), 
    is.na(event_count_lact) ~ 'Unknown',
    TRUE~'Mast 4+'
  ))%>%
  
  #match the dates to the denominator
  mutate(date_time_period_start = floor_date(date_event, unit = 'month'))%>% 
  
  #use same  lactation grouping as denominator (0, 1, 2, 3+), 
  #and set location to match denominator (locations will depend of farm data structure)
  mutate(`Lactation Group` = lact_group, 
         location_lact_list = location_event) %>% 
  
  #group by all variables in the denominator, plus any extras that you want (like event count), and summarize
  group_by(location_lact_list, `Lactation Group`, 
            date_time_period_start, event_count_lact)%>% 
  summarize(ct_mast_events = n_distinct(id_animal))%>%
  ungroup()
  

```

### Step 3 - Join denominator to numerator

Be careful with these joins.  How do you know it joined correctly?
Add some code to this chunk to demonstrate that it performed the way you expected.

```{r, echo = TRUE}
final_data<-denominator%>%
  left_join(mast_event_summary)%>%
  mutate(ct_mast_events = replace_na(ct_mast_events, 0))%>%
  mutate(mast_rate = ct_mast_events/ct_animal_time_periods)

head(final_data)
```


### Step 4 - Look at results

```{r, echo = TRUE}

ggplot(final_data%>%filter(!(is.na(event_count_lact))))+
  geom_point(aes(x = date_time_period_start, y= mast_rate, color = `Lactation Group`))+
    geom_smooth(aes(x = date_time_period_start, y= mast_rate, 
                    color = `Lactation Group`, group =  `Lactation Group`, 
                    fill = `Lactation Group`),alpha = .2 , span = .6)+

  facet_grid(event_count_lact~location_lact_list, scales = 'free')



```
## Step 5 - your turn

Repeat the steps above for a disease of your choice. Then plot the trend of that disease incidence over time.

Be careful with your joins. Do not unintentionally create or delete data.

```{r}
# =========================================================================
# STEP 5: DISEASE INCIDENCE ANALYSIS (RETAINED PLACENTA - RP)
# =========================================================================

# -------------------------------------------------------------------------
# PART 1: Summarize the Numerator (RP Events)
# -------------------------------------------------------------------------
rp_event_summary <- read_parquet(here::here('data/intermediate_files/events_formatted.parquet')) %>%
  
  # Filter for the specific disease code 'RP' (Retained Placenta)
  filter(event %in% 'RP') %>% 
  
  # REMOVE DUPLICATES: This prevents the 'must be size 0 or 1, not 2' error in the professor's function
  distinct(id_animal, date_event, .keep_all = TRUE) %>%
  
  # Execute the custom function to calculate historical event counts per cow
  fxn_add_event_counts() %>%  
  
  # Standardize labels and group higher occurrence frequencies together
  mutate(event_count_lact = case_when(
    event_count_lact < 4    ~ paste0('RP ', event_count_lact), 
    is.na(event_count_lact) ~ 'Unknown',
    TRUE                    ~ 'RP 4+'
  )) %>%
  
  # Roll back the exact event dates to the first day of their respective month
  mutate(date_time_period_start = floor_date(date_event, unit = 'month')) %>% 
  
  # Align column naming conventions directly with the denominator dataset
  mutate(`Lactation Group` = lact_group, 
         location_lact_list = location_event) %>% 
  
  # Group the rows by structural dimensions before evaluating aggregates
  group_by(location_lact_list, `Lactation Group`, 
           date_time_period_start, event_count_lact) %>% 
  
  # Count unique animal identification keys inside each isolated group
  summarize(ct_rp_events = n_distinct(id_animal)) %>%
  
  # Strip grouping attributes so the dataframe functions normally downstream
  ungroup()


# -------------------------------------------------------------------------
# PART 2: Combine Denominator and Numerator (Left Join)
# -------------------------------------------------------------------------
final_rp_data <- denominator %>%
  
  # Merge data arrays horizontally using matching relational key columns
  left_join(rp_event_summary) %>%
  
  # Replace structural NA voids with 0 for groups with no recorded events
  mutate(ct_rp_events = replace_na(ct_rp_events, 0)) %>%
  
  # Calculate the population incidence scale (infected count over active population)
  mutate(rp_rate = ct_rp_events / ct_animal_time_periods)


# -------------------------------------------------------------------------
# PART 3: Generate Trend Visualization (ggplot2)
# -------------------------------------------------------------------------
ggplot(final_rp_data %>% filter(!(is.na(event_count_lact)))) +
  
  # Render absolute mathematical coordinate points onto the visual canvas
  geom_point(aes(x = date_time_period_start, y = rp_rate, color = `Lactation Group`)) +
  
  # Overlay a mathematical smoothing path to illustrate directional trends
  geom_smooth(aes(x = date_time_period_start, y = rp_rate, 
                  color = `Lactation Group`, group = `Lactation Group`, 
                  fill = `Lactation Group`), alpha = .2 , span = .6) +
  
  # Segment the graphics display matrix by event categories and farm boundaries
  facet_grid(event_count_lact ~ location_lact_list, scales = 'free')
```






## Part 2 - Extension

Think about your disease or event of interest. Standard denominators like the one above are often not good enough.

Use your new data wrangling skills to create the perfect denominator for your question.


NOTE: Denominators can be challenging. If you get stuck:

-   First articulate what you want to do, step by step, in English

-   Then add a blank code chunk for each step

-   Begin filling in code

-   Ask the group for help if you get completely stuck.
