---
title: "Week 4 tidy and join"
format:
  html:
    embed-resources: true
    code-fold: true
---

# ignore this chunk, it is just alot of setup

```{r, include = FALSE}

library(tidyverse)
source('functions/fxn_mileston4_tools.R')
```

# Week 4

## FDA facilities

The object `fda_facilities_solution_data` is a dataframe of the number of fda approved facilities in the world. The original data can be found at the link below.

<https://www.fda.gov/economics-staff/fda-glance>

Your ultimate goal is to create this graphic:

```{r}

fxn_plot_facilities(fda_facilities_solution_data)

```

-   Make the data in the chunk below tidy.

-   Calculate the percent of the total for each year that is domestic or foreign.

Your final data in this chunk should be named `fda_facilities`, and should look like the dataframe in your environment named `fda_facilities_solution_data`.
# | Tidying the data | #
```{r}
fda_facilities<-tibble(
  Year = c(2018, 2020, 2021, 2023, 2024),
  Domestic = c(1403, 1608, 1858, 980, 974), 
  Foreign = c( 385, 1156, 1103, 701, 753),
  Total = c(1788, 2764, 2961, 1681, 1727)
) |>
  pivot_longer(cols = 2:3,
  names_to = "Location",
  values_to = "Facility Count") |>
    mutate(pct = `Facility Count` / Total) #have to put Facility Count in backticks due to space#


```

In the chunk below:

-   create a bar graph with year on the x axis, a count of fda facilities on the y axis, color the bar by location.

-   use geom text to add text to your graphic showing the total number of facilities, the percent foreign and percent domestic. Use the graphic in chunk 2 as an example of what you are trying to create.

# | Facility count plots | #
```{r}
fda_facilities_plot1 <- fda_facilities |>
  ggplot(mapping = aes(x = factor(Year))) +                # Used factor(Year) to treat as categories so all years would be listed
  geom_col(data = distinct(fda_facilities, Year, Total),   # Needed this to not count Total twice
  mapping = aes(y = Total),
  fill = "#36454F",
  width = 0.9) +                                           # Made the Total bar wider than the Location bars 
  geom_col(mapping = aes(y = `Facility Count`, fill = Location),
  position = "dodge",
  width = 0.7) +                                           # Made the Location bars smaller
    scale_fill_manual(                                     # scale_fill_brewer(name = "Location", palette = "Blues")
      values = c(
    Domestic = "#45a29e",
    Foreign  = "#ffa552"
  )
    ) +
  labs(
    title = "FDA Facilities - Animal Drugs",
    subtitle = "Numbers on grey total bars are facility counts (% domestic)",
    x = "Year",
    y = "Total",
    caption = "Data from https://www.fda.gov/economics-staff/fda-glance"
  ) +
  geom_text(
    data = filter(fda_facilities, Location == "Domestic"), # Filtering the data so that I only pull percent domestic and only pull Total from each year once
    aes(
      y = Total,
      label = paste0(
        Total,
        "\n",
        sprintf("(%d%%)", round(pct * 100))                # Used this format just to try something new. Could have also written as "\n(", round(pct * 100), "%)"
      )),
      color = "white",
    vjust = 1.3
  ) +
    theme_minimal()

  fda_facilities_plot1
  ggsave("milestones_dairy/images/fda_facilities_plot1.png", plot = fda_facilities_plot1)
  

```



## FDA Greenbook

All ***listed*** animal drug products (which is not the same as ***approved*** products) can be found in the FDA "Green Book".

Go to <https://www.fda.gov/industry/structured-product-labeling-resources/electronic-animal-drug-product-listing-directory>

Then click on the link text at the top of the page that says `Electronic Animal Drug Product Listing Directory`. This will download an excel file (but a very very old version of excel) to your computer.

Put this file in a place you can then read it into R in the chunk below. Make sure when you read it in that you force it to read in all columns as characters. It will make errors if it guesses on this dataset.

Hint: you can either learn to read in very old excel files by finding a function that (readxl), or you can open the file and re-save it as a csv file and then read it in (potentially faster)

Pipe your data into the function `fxn_clean_green_manufacturer()` to clean up the manufacturer names.  Group by manufacturer and count how many products are manufactured. Arrange the data by descending number of products. 

Make a new dataframe named `Top_10` that is only the first 10 rows.

# Reading in the Greenbook
```{r}
fda_greenbook <- read.csv("C:/Users/peppe/Documents/Posit course/Electronic_Animal_Drug_Product_Listing_Directory/Electronic_Animal_Drug_Product_Listing_Directory.xls.csv")

install.packages("skimr")
library(skimr)
skim(fda_greenbook)
summary(fda_greenbook, character.method = "factor")

```


Now join this data to the `pharma_revenue` data frame that is already in your environment.  



```{r}
glimpse(fda_greenbook)
glimpse(pharma_revenue)
pharma_revenue |>
  head(13)

fda_greenbook |>
  filter(
    str_detect(
      str_to_lower(Labeler.Name),
      "zoetis|elanco|merck|dechra|boehringer|bimeda|mwi|vetone|norbrook|phibro|virbac"
    )
  ) |>
  distinct(Labeler.Name) |>
  arrange(Labeler.Name)

# Have to recode Labeler.Name from fda_greenbook, since manufacturer names are in multiple formats
# If I had read the instructions, I wouldn't have needed to do that...
fda_greenbook_1 <- fda_greenbook |>
  mutate(
    Manufacturer = case_when(
      str_detect(Labeler.Name, regex("bimeda", ignore_case = TRUE)) ~ "Bimeda, Inc.",
      str_detect(Labeler.Name, regex("boehringer", ignore_case = TRUE)) ~ "Boehringer",
      str_detect(Labeler.Name, regex("dechra", ignore_case = TRUE)) ~ "Dechra",
      str_detect(Labeler.Name, regex("elanco", ignore_case = TRUE)) ~ "Elanco",
      str_detect(Labeler.Name, regex("merck|intervet", ignore_case = TRUE)) ~ "Merck",
      str_detect(Labeler.Name, regex("mwi|vetone|vet\\s*one", ignore_case = TRUE)) ~ "MWI/VetOne",
      str_detect(Labeler.Name, regex("norbrook", ignore_case = TRUE)) ~ "Norbrook Laboratories Limited",
      str_detect(Labeler.Name, regex("phibro", ignore_case = TRUE)) ~ "Phibro Animal Health",
      str_detect(Labeler.Name, regex("virbac", ignore_case = TRUE)) ~ "Virbac",
      str_detect(Labeler.Name, regex("zoetis", ignore_case = TRUE)) ~ "Zoetis",
      TRUE ~ NA_character_
    )
  )

pharma_revenue_clean <- pharma_revenue |> # Remove duplicates to avoid 'many-to-many' relationships warning and multiplying rows #
  distinct() |>
  mutate(
    `Revenue (millions)` = parse_number(`Gross Revenue (Animal Health Only)`) *
    case_when(
  str_detect(`Gross Revenue (Animal Health Only)`, "Billion") ~ 1000,
  str_detect(`Gross Revenue (Animal Health Only)`, "Million") ~ 1
    )
  )

greenbook_revenue <- inner_join(pharma_revenue_clean, fda_greenbook_1, by = "Manufacturer")
greenbook_rev_left <- left_join(pharma_revenue_clean, fda_greenbook_1, by = "Manufacturer")

# Summarizing by manufacturer, revenue, number of NDC to see if there is a correlation between revenue and number of listed products
summary_greenbook_rev <- greenbook_rev_left |>
  group_by(Manufacturer, `Revenue (millions)`) |>
  summarize(
    sum(n())
  ) |>
    arrange(desc(`Revenue (millions)`))

  summary_greenbook_rev |>
    rename(`Number of products` = `sum(n())`) |>
    ggplot() +
  geom_point(mapping = aes(x = `Number of products`, y = `Revenue (millions)`, color = Manufacturer), size = 4) +
  scale_color_brewer(name = "Manufacturer", palette = "Paired") +
  theme_minimal()
```


Did it join as expected?

Is there any correlation between number of products listed and revenue?

## Extension

Explore something else in this greenbook data. Can you summarize 2 different things and then join them back on a shared variable?

### Is there a correlation between list type and revenue?
```{r}
# See if there is a difference between revenue and number of NADA, ANADA, or other categories (Marketing Category variable) and product type (OTC, Rx, VFD, bulk, other) and revenue and join back on revenue

# Create dataframe with major manufacturers and summary of revenue vs. marketing category
greenbook_rev_market <- greenbook_rev_left |>
  select(Manufacturer, `Revenue (millions)`, NDC, Application.Number.or.Monograph.ID, Marketing.Category, Proprietary.Name, Non.Proprietary.Name, Ingredient.List) |>
  mutate(
    Approval.type = case_when(
      str_detect(Marketing.Category, regex("anada", ignore_case = TRUE)) ~ "ANADA",
      str_detect(Marketing.Category, regex("conditional", ignore_case = TRUE)) ~ "Conditional",
      str_detect(Marketing.Category, regex("nada", ignore_case = TRUE)) ~ "NADA",
      str_detect(Marketing.Category, regex("minor", ignore_case = TRUE)) ~ "Indexed",
      TRUE ~ "Other unapproved"
    )
  )

summary_greenbook_market <- greenbook_rev_market |>
  add_count(Manufacturer, `Approval.type`, name = "n_products")
  
#greenbook_market_plot <- summary_greenbook_market |>
#  ggplot() +
#  facet_wrap(facets = vars(Approval.type), scales = "fixed", nrow = 1) +
#  geom_point(
#    mapping = aes(x = n_products, y = `Revenue (millions)`, color = Manufacturer),
#    size = 3.5,
#    alpha = 0.6
#  ) +
#  geom_smooth(
#    mapping = aes(x = n_products, y = `Revenue (millions)`),
#    method = "lm",
#    se = FALSE,
#    color = "black",
#    linewidth = 0.7
#  ) +
#  scale_color_brewer(name = "Manufacturer", palette = "Paired") +
#  labs(
#    x = "Number of products",
#    title = "Greenbook products by marketing category and manufacturer"
#  ) +
#  theme_minimal()
#
#greenbook_market_plot

## This is the best one. Used Claude to play with different plot types and help me find a colorblind friendly palette.
greenbook_market_nofacet_plot <- summary_greenbook_market |>
  ggplot() +
  geom_point(
    mapping = aes(x = n_products, y = `Revenue (millions)`, shape = Manufacturer, color = Approval.type),
    size = 3.5,
    alpha = 0.6
  ) +
  geom_smooth(
    mapping = aes(x = n_products, y = `Revenue (millions)`, color = Approval.type),
    method = "lm",
    se = TRUE,
    linewidth = 0.7
  ) +
  scale_shape_manual(values = 0:9) +
  scale_color_manual(
    values = c(
      "ANADA" = "#2a78d6",
      "Conditional" = "#008300",
      "NADA" = "#e87ba4",
      "Indexed" = "#eda100",
      "Other unapproved" = "#898781"
    )
  ) +
  labs(
    x = "Number of products",
    color = "Approval type",
    shape = "Manufacturer",
    title = "Greenbook revenue by number of products, approval type, and manufacturer"
  ) +
  theme_minimal()

greenbook_market_nofacet_plot

#greenbook_market_strip_plot <- summary_greenbook_market |>
#  ggplot(mapping = aes(x = Approval.type, y = `Revenue (millions)`, color = Manufacturer)) +
#  geom_jitter(width = 0.2, size = 3.5, alpha = 0.6) +
#  scale_color_brewer(name = "Manufacturer", palette = "Paired") +
#  labs(
#    x = "Approval type",
#    title = "Greenbook revenue by marketing category and manufacturer"
#  ) +
#  theme_minimal()
#
#greenbook_market_strip_plot

#greenbook_market_bubble_plot <- summary_greenbook_market |>
#  ggplot(mapping = aes(x = Approval.type, y = `Revenue (millions)`, size = n_products, color = Manufacturer)) +
#  geom_jitter(width = 0.2, alpha = 0.6) +
#  scale_color_brewer(name = "Manufacturer", palette = "Paired") +
#  scale_size_continuous(name = "Number of products") +
#  labs(
#    x = "Approval type",
#    title = "Greenbook revenue by marketing category and manufacturer"
#  ) +
#  theme_minimal()
#
#greenbook_market_bubble_plot


# Create dataframe with major manufacturers and summary of revenue vs. product type
greenbook_rev_product <- greenbook_rev_left |>
  select(Manufacturer, `Revenue (millions)`, NDC, Application.Number.or.Monograph.ID, Product.Type, Proprietary.Name) |>
  mutate(
    Drug.type = case_when(
      str_detect(Product.Type, regex("vfd", ignore_case = TRUE)) ~ "VFD Type A",
      str_detect(Product.Type, regex("article", ignore_case = TRUE)) ~ "OTC Type A",
      str_detect(Product.Type, regex("prescription", ignore_case = TRUE)) ~ "Rx",
      str_detect(Product.Type, regex("otc", ignore_case = TRUE)) ~ "OTC",
      str_detect(Product.Type, regex("bulk", ignore_case = TRUE)) ~ "Bulk",
      TRUE ~ "Other"
    )
  ) |>
    relocate(Drug.type, .before = Product.Type)

summary_greenbook_product <- greenbook_rev_product |>
  add_count(Manufacturer, `Drug.type`, name = "n_drug")


greenbook_product_plot <- summary_greenbook_product |>
  ggplot() +
  geom_point(
    mapping = aes(x = n_drug, y = `Revenue (millions)`, shape = Manufacturer, color = Drug.type),
    size = 3.5,
    alpha = 0.6
  ) +
  geom_smooth(
    mapping = aes(x = n_drug, y = `Revenue (millions)`, color = Drug.type),
    method = "lm",
    se = TRUE,
    linewidth = 0.7
  ) +
  scale_shape_manual(values = 0:9) +
  scale_color_manual(
    values = c(
      "VFD Type A" = "#2a78d6",
      "OTC Type A" = "#008300",
      "Rx" = "#e87ba4",
      "OTC" = "#eda100",
      "Bulk" = "#898781"
    )
  ) +
  labs(
    x = "Number of products",
    color = "Drug type",
    shape = "Manufacturer",
    title = "Greenbook revenue by number of products, drug type, and manufacturer"
  ) +
  theme_minimal()

greenbook_product_plot

# Try to see individual relationships with facets
## Rx appear to have strongest correlation to revenue
greenbook_product_plot_1 <- summary_greenbook_product |>
  ggplot() +
  facet_wrap(facets = vars(Drug.type), scales = "fixed", nrow = 1) +
  geom_point(
    mapping = aes(x = n_drug, y = `Revenue (millions)`, color = Manufacturer),
    size = 3.5,
    alpha = 0.6
  ) +
  geom_smooth(
    mapping = aes(x = n_drug, y = `Revenue (millions)`),
    method = "lm",
    se = FALSE,
    color = "black",
    linewidth = 0.7
  ) +
  scale_color_brewer(name = "Manufacturer", palette = "Paired") +
  labs(
    x = "Number of products",
    title = "Greenbook products by drug category and manufacturer"
  ) +
  theme_minimal()

greenbook_product_plot_1

# Join marketing and product type dataframes back on manufacturer
summary_fda_drugcat_rev <- left_join(summary_greenbook_product, summary_greenbook_market, join_by(Manufacturer, `Revenue (millions)`, NDC, Application.Number.or.Monograph.ID, Proprietary.Name))
```

# Week 5

## Clean up Marketing Category

Use your new skills for working with strings to clean up the `Marketing Category` column in the greenbook.

```{r}
greenbook_rev_left |>
  skim() # There are 12 unique marketing categories

greenbook_rev_left |>
  group_by(Manufacturer, Marketing.Category) |>
  summarize(
    sum(n())
  ) |>
    arrange(desc(`Manufacturer`)) |>
    print(n = 44)

# Would like to keep NADA, ANADA, indexed (Index), conditional (CA), bulk ingredient (Bulk), export (Exp), other unapproved (Unapp)
## NADA and ANADA don't need to be cleaned
### Already did this in week 4 extension
greenbook_clean <- fda_greenbook |>
  mutate(
    Approval.type = case_when(
      str_detect(Marketing.Category, regex("anada", ignore_case = TRUE)) ~ "ANADA",
      str_detect(Marketing.Category, regex("conditional", ignore_case = TRUE)) ~ "Conditional",
      str_detect(Marketing.Category, regex("nada", ignore_case = TRUE)) ~ "NADA",
      str_detect(Marketing.Category, regex("minor", ignore_case = TRUE)) ~ "Indexed",
      TRUE ~ "Other unapproved"
    )
  ) |>
    relocate(Approval.type, .before = Proprietary.Name)
```


## Format Dates

Use your date formating skills to format `Inactivation Date` and `Reactivation Date`.  

```{r}
greenbook_clean <- greenbook_clean|>
  mutate(
    Inactivation.Date = ymd(Inactivation.Date),
    Reactivation.Date = ymd(Reactivation.Date)
  )

```


Is there a particular month of the year when more approvals are inactivated?

```{r}

```


## Extension

Explore the greenbook using these new varaibles and anything else you are interested in (such as Ingredient List?).  You may find you also wish to clean up the ingredient list if you choose to use it.

```{r}

```


