---
title: "Milestone 4 - Denominators (Tidy and Join)"
format: html
date: "`r Sys.Date()`"
embed-resources: true
---

```{r}
library(tidyverse)
library(quarto)
library(arrow)
library(waldo)

source('functions/fxn_add_event_counts.R')

solution<-read_rds(here::here('milestones_dairy/data_milestones/solutions_week_05_pivotlonger.rds'))

```

## Denominators

Now that you can tidy data, pivot longer or wider, and then join tables to others, you are both unstoppable . . . and dangerous.

In this lesson we will explore one method of using very granular denominators in your reports.

We begin by reading in files for events, and denominators

```{r initial denominator generation, message==FALSE}

denominator_base<-read_parquet(here::here('data/intermediate_files/denominator_by_calendar_time_period.parquet'))
events_formatted <- read_parquet(here::here('data/intermediate_files/events_formatted.parquet'))

```


The first step is to filter your denominator according to the `calendar_time_period_type` that you want. 

The options for calendar periods are: **`r sort(unique(denominator_base$calendar_time_period_type))`**

The second step is to choose your lactation grouping. 
To simplify this first example will select a grouping without day of phase. 
(see the DOP section of How to Use denominaotrs report for day of phase examples)

The options for groups are: **`r sort(unique(denominator_base$deno_type))`**

Your final code should look something like this:

```{r, echo = TRUE}

denominator<-read_parquet(here::here('data/intermediate_files/denominator_by_calendar_time_period.parquet'))%>%
  filter(calendar_time_period_type %in% 'month')%>% #choose denominator period type
  filter(deno_type %in% 'lact_3+')%>% #choose the type of lactation group
  distinct()%>%
  select(-contains('dop'), -day_of_phase_group) # day of phase columns are not needed when dop is not part of grouping

```

```{r}
head(denominator)
```
### Step 2 - Summarize your numerator correctly

Follow the comments documenting the steps for nummerator summary. 

Make sure you understand them.  This is a functional example of what you will be asked to do in step 5 for an event or disease of your choice.


```{r, echo = TRUE}

mast_event_summary<-read_parquet(here::here('data/intermediate_files/events_formatted.parquet'))%>%
  
  #select event of interest
  filter(event %in% 'MAST')%>% #
  
  #add the event counts the way you like
  fxn_add_event_counts()%>% 
  mutate(event_count_lact = case_when(
    event_count_lact <4~paste0('Mast ', event_count_lact), 
    is.na(event_count_lact) ~ 'Unknown',
    TRUE~'Mast 4+'
  ))%>%
  
  #match the dates to the denominator
  mutate(date_time_period_start = floor_date(date_event, unit = 'month'))%>% 
  
  #use same  lactation grouping as denominator (0, 1, 2, 3+), 
  #and set location to match denominator (locations will depend of farm data structure)
  mutate(`Lactation Group` = lact_group, 
         location_lact_list = location_event) %>% 
  
  #group by all variables in the denominator, plus any extras that you want (like event count), and summarize
  group_by(location_lact_list, `Lactation Group`, 
            date_time_period_start, event_count_lact)%>% 
  summarize(ct_mast_events = n_distinct(id_animal))%>%
  ungroup()
  

```

### Step 3 - Join denominator to numerator

Be careful with these joins.  How do you know it joined correctly?
Add some code to this chunk to demonstrate that it performed the way you expected.

```{r, echo = TRUE}
final_data<-denominator%>%
  left_join(mast_event_summary)%>%
  mutate(ct_mast_events = replace_na(ct_mast_events, 0))%>%
  mutate(mast_rate = ct_mast_events/ct_animal_time_periods)

head(final_data)
```


### Step 4 - Look at results

```{r, echo = TRUE}

ggplot(final_data%>%filter(!(is.na(event_count_lact))))+
  geom_point(aes(x = date_time_period_start, y= mast_rate, color = `Lactation Group`))+
    geom_smooth(aes(x = date_time_period_start, y= mast_rate, 
                    color = `Lactation Group`, group =  `Lactation Group`, 
                    fill = `Lactation Group`),alpha = .2 , span = .6)+

  facet_grid(event_count_lact~location_lact_list, scales = 'free')



```
## Step 5 - your turn

Repeat the steps above for a disease of your choice. Then plot the trend of that disease incidence over time.

Be careful with your joins. Do not unintentionally create or delete data.

```{r}
#Great, I want to look at aborts in pregnant/dry cows by month
#I'm assuming the RC column in events_formatted is DC305 Repro Code, so I'll use cows with RC==5,6 as a denominator
#are there multiple locations in this dataset?
events_formatted|>
  distinct(location_event)|>
  arrange(location_event)
#lol nope. Just Example herd 210 so we can ignore location.

#So pregnant & dry cows in the denominator separated out by month
preg_denominator <- events_formatted |>
  filter(RC %in% c("5", "6")) |> #find only cows with RC==5,6
  mutate(
    date_time_period_start = floor_date(date_event, unit = "month"), #reframes dates as the whole month (so 1/12/25 = january 2025)
  ) |>
  group_by(date_time_period_start) |> #then group by month
  summarize(
    ct_preg_animals = n_distinct(id_animal), #count the pregnant animals (by distinct ID) in each month
    .groups = "drop"
  )

#my numerator is the abortions by month
abort_event_summary<- events_formatted|>
  filter(event=="ABORT")|>
  mutate(
    date_time_period_start = floor_date(date_event, unit="month")
  )|>
  group_by(date_time_period_start)|>
  summarize(ct_abort_events = n_distinct(id_animal), #same as above, just looking at abort events rather than rc=5,6
  .groups="drop")

#now to join them
final_abort_over_preg <-preg_denominator |> #start with denominator
  left_join(abort_event_summary, by = "date_time_period_start")|> #join datasets by month
  mutate(ct_abort_events = replace_na(ct_abort_events, 0), #turn any missing abort events to "0" so we can math it
  abort_rate_preg = ct_abort_events/ct_preg_animals) #puts the numerator over denominator to calculate abortion rate

#now let's graph by month!
ggplot(final_abort_over_preg) +
  geom_point(aes(x=date_time_period_start, y=abort_rate_preg))+
  geom_smooth(aes(x=date_time_period_start, y=abort_rate_preg), span=0.6, alpha=0.2)+
  scale_y_continuous(labels = scales::percent)+ #fixes y axis to show percentages instead of decimals as calculated in previous step
  labs(
    title="Abortion rate by month",
    x="Month",
    y="Abort rate")

#this is fine, but I want to break out the months more - ChatGPT helped me with this.
ggplot(final_abort_over_preg) +
  geom_point(aes(x = date_time_period_start,y = abort_rate_preg))+
  geom_line(aes(x = date_time_period_start, y = abort_rate_preg)) +
  scale_x_date(date_breaks = "1 month",date_labels = "%b %Y") + #this is how I'll break out the months with month (abbreviated), year
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Abortion rate by month",
    x = "Month",
    y = "Abort rate"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) #adjusts labels on x axis to a 45 degree angle so they're not overlapping each other & align right so label aligns with tick mark better

#okay that looks legit crazy... let's look at just the last year
final_abort_over_preg |>
  filter(date_time_period_start >= max(date_time_period_start, na.rm = TRUE) - years(1)) |> #just hte last year from teh most recent date in teh dataset
  ggplot() +
  geom_col(aes(x = date_time_period_start,y = abort_rate_preg)) +
  geom_smooth(
    aes(x = date_time_period_start, y = abort_rate_preg),span = 0.6,alpha = 0.2) +
  scale_x_date(date_breaks = "1 month",date_labels = "%B %Y") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Abortion rate by month: Last year",
    x = "Month",
    y = "Abort rate"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#much more readable. I like the columns in this context. something happened in march!
```






## Part 2 - Extension

Think about your disease or event of interest. Standard denominators like the one above are often not good enough.

Use your new data wrangling skills to create the perfect denominator for your question.


NOTE: Denominators can be challenging. If you get stuck:

-   First articulate what you want to do, step by step, in English

-   Then add a blank code chunk for each step

-   Begin filling in code

-   Ask the group for help if you get completely stuck.
