---
title: "Milestone 5 - Dates and Strings"
format: html
date: "`r Sys.Date()`"
embed-resources: true
---

```{r}
library(tidyverse)
library(quarto)
library(arrow)
library(waldo)

solution<-read_parquet('data/intermediate_files/events_formatted.parquet')

# read in files-----------------

list_files <- list.files(here::here("data/event_files")) # folder name where event files are located

events_original <- NULL

# i=1
for (i in seq_along(list_files)) {
  df <- read_csv(
    here::here(
      "data/event_files",
      list_files[i]
    ),
    # reads in all data as character string
    col_types = cols(.default = "c")
  )
  
   events_original <- bind_rows(events_original, df)
}

#holding point for me to come back to

```

## Dates

The original dairy comp data is read into your environment and named "events_original".
For consistency all columns are read in as characters.

Use your new skills formating dates to create a new column named date_event that is a real date from the original column named Date.

Then create new variables that provide floor dates by month, season, and year. Name these floordate_month, floordate_year, floordate_season.

```{r initial denominator generation, message==FALSE}
events_original<-events_original |>
  mutate(
    date_event=mdy(Date),
    floordate_month=floor_date(date_event, "month"),
    floordate_season=floor_date(date_event, "season"),
    floordate_year=floor_date(date_event, "year")
  )

events_original #checking to see tibble is what I need with new variables and corrected characters to dates

```

Do your new dates match the dates in the solution?


# Strings

Find all the columns in the solution that contain the word "remark".  

These are all based on the original Remark column in dairy comp. 

Can you recreate them?  

Hint: don't spend too much time trying to get exactly the same values. 
Just spend time thinking about how you could use them to clean up remarks in a way that would help you answer a question.

## My Questions about Extract All issues?
```{r, echo = TRUE}

 #unsure why extract all isn't giving me the data in a readable format so I can split it up...is it because it is not tidy data and generating rows of variables vs columns of variables?
  # My from pen to pen move example that does not work
  # str_extract_all(Remark,"F(\\d{3})T(\\d{3})") #extract numbers between character letters...ie F201T202 gives "201"


events_original<-events_original |>
 
  mutate(
    remark_numbers1=str_extract(Remark,"\\d+"),
    remark_letters1=str_extract(Remark,"[A-Za-z]+"), #this will extract the first letters in a character string and ignore numbers
    remark_remaining_after_numbers1=str_remove(Remark,"\\d+"),#has leftovers (numbers and letters) after numbers1 like "FT###"
    remark_remaining_after_letters1=str_remove(Remark,"[A-Za-z]+"),#has leftovers (numbers and letters) after letters1 like "###T###"
    remark_numbers2=str_extract(remark_remaining_after_letters1,"\\d+"),  #OR could do
        #remark_numbers2=str_extract(Remark,"\\d+(?!.*\\d+)"), #not sure how to do this in a simpler way, this looks backwards for numerics in character strings
    remark_letters2=str_extract(remark_remaining_after_numbers1,"[A-Za-z]+"), #looks like its repeating remark_letters1 but using remark_remaining_after_numbers1 as data frame
    
    extracted_movefrom_pen=str_extract(Remark,"(?<=F)\\d{3}"), #added pen move extractions here...might be better ways
    extracted_moveto_pen=str_extract(Remark,"(?<=T)\\d{3}")
  )

events_original



```

# Extension - Your Turn

Start with events_original.  

Use your wrangling skils to fix date formats, and simplify free text to answer a question on this dairy.

You can choose any question you want. 

Here are some examples:

-  Which drug is most commonly used for mastitis treatment and has this changed over time?
-  What percent of breedings use sexed semen?  (this requires a large wrangle with an external key)
-  What specific lameness diagnosis are there and how frequently do they occur? 

```{r, echo = TRUE}

```
