---
title: "Milestone 4 - Denominators (Tidy and Join)"
format: html
date: "`r Sys.Date()`"
embed-resources: true
---

```{r}
library(tidyverse)
library(quarto)
library(arrow)
library(waldo)

solution<-read_rds(here::here('milestones_dairy/data_milestones/solutions_week_05_pivotlonger.rds'))

```

## Denominators

Now that you can tidy data, pivot longer or wider, and then join tables to others, you are both unstoppable . . . and dangerous.

In this lesson we will explore one method of using very granular denominators in your reports.

We begin by reading in files for events, and denominators

```{r initial denominator generation, message==FALSE}

denominator_base<-read_parquet(here::here('data/intermediate_files/denominator_by_calendar_time_period.parquet'))
events_formatted <- read_parquet(here::here('data/intermediate_files/events_formatted.parquet'))

```


The first step is to filter your denominator according to the `calendar_time_period_type` that you want. 

The options for calendar periods are: **`r sort(unique(denominator_base$calendar_time_period_type))`**

The second step is to choose your lactation grouping. 
To simplify this first example will select a grouping without day of phase. 
(see the DOP section of How to Use denominaotrs report for day of phase examples)

The options for groups are: **`r sort(unique(denominator_base$deno_type))`**

Your final code should look something like this:

```{r, echo = TRUE}

denominator<-read_parquet(here::here('data/intermediate_files/denominator_by_calendar_time_period.parquet'))%>%
  filter(calendar_time_period_type %in% 'month')%>% #choose denominator period type
  filter(deno_type %in% 'lact_3+')%>% #choose the type of lactation group
  distinct()%>%
  select(-contains('dop'), -day_of_phase_group) # day of phase columns are not needed when dop is not part of grouping

denominator

```

```{r}
head(denominator)
```
### Step 2 - Summarize your numerator correctly

Follow the comments documenting the steps for nummerator summary. 

Make sure you understand them.  This is a functional example of what you will be asked to do in step 5 for an event or disease of your choice.


```{r, echo = TRUE}

mast_event_summary<-read_parquet(here::here('data/intermediate_files/events_formatted.parquet'))%>%
  
  #select event of interest
  filter(event %in% 'MAST')%>% #
  
  #add the event counts the way you like
  fxn_add_event_counts()%>% 
  mutate(event_count_lact = case_when(
    event_count_lact <4~paste0('Mast ', event_count_lact), 
    is.na(event_count_lact) ~ 'Unknown',
    TRUE~'Mast 4+'
  ))%>%
  
  #match the dates to the denominator
  mutate(date_time_period_start = floor_date(date_event, unit = 'month'))%>% 
  
  #use same  lactation grouping as denominator (0, 1, 2, 3+), 
  #and set location to match denominator (locations will depend of farm data structure)
  mutate(`Lactation Group` = lact_group, 
         location_lact_list = location_event) %>% 
  
  #group by all variables in the denominator, plus any extras that you want (like event count), and summarize
  group_by(location_lact_list, `Lactation Group`, 
            date_time_period_start, event_count_lact)%>% 
  summarize(ct_mast_events = n_distinct(id_animal))%>%
  ungroup()
  

```

### Step 3 - Join denominator to numerator

Be careful with these joins.  How do you know it joined correctly?
Add some code to this chunk to demonstrate that it performed the way you expected.

```{r, echo = TRUE}
final_data<-denominator%>%
  left_join(mast_event_summary)%>%
  mutate(ct_mast_events = replace_na(ct_mast_events, 0))%>%
  mutate(mast_rate = ct_mast_events/ct_animal_time_periods)

head(final_data)
```


### Step 4 - Look at results

```{r, echo = TRUE}

ggplot(final_data%>%filter(!(is.na(event_count_lact))))+
  geom_point(aes(x = date_time_period_start, y= mast_rate, color = `Lactation Group`))+
    geom_smooth(aes(x = date_time_period_start, y= mast_rate, 
                    color = `Lactation Group`, group =  `Lactation Group`, 
                    fill = `Lactation Group`),alpha = .2 , span = .6)+

  facet_grid(event_count_lact~location_lact_list, scales = 'free')



```
## Step 5 - your turn

Repeat the steps above for a disease of your choice. Then plot the trend of that disease incidence over time.

Be careful with your joins. Do not unintentionally create or delete data.

```{r}

#Effect of early calfhood disease on "stayability"

library(survival)
library(survminer)

events_formatted
events_formatted$event
events_formatted |> filter(event_type == "health") |> count(event, sort=TRUE)

all_animals <- events_formatted |> distinct(id)
all_animals

calf_health <- events_formatted |> filter(age_event <= 30, event %in% c("SCOURS","PNEU")) |> 
  group_by(id) |>
  summarize(
    scours = any(event == "SCOURS"), 
    pneu = any(event == "PNEU")) |>
              mutate(disease_group = case_when(scours & pneu ~ "Both",
                                               scours ~ "Scours",
                                               pneu ~ "Pneu",
                                               TRUE ~ "None"))
            
  calf_health          

  analysis_calfhealth <- all_animals |> left_join(calf_health, by = "id") |>
  mutate(
    scours = replace_na(scours,FALSE),
    pneu = replace_na(pneu,FALSE),
    disease_group = case_when(scours & pneu ~ "Both",
                              scours ~ "Scours",
                              pneu ~ "Pneu",
                              TRUE ~ "None"))


analysis_calfhealth

#get sold events for animals

sold_dates <- events_formatted |>
  filter(event == "SOLD") |>
  group_by(id) |>
  summarize(
    date_sold = min(date_event),
    .groups = "drop"
  )

#get a start date - from first freshening

fresh_date <- events_formatted |> 
  filter(event == "FRESH") |>
  group_by(id) |>
  summarize(
    date_fresh = min(date_event),
    .groups = "drop"
  )

#add died... no died events??

final_calfhealth <- analysis_calfhealth |>
  left_join(sold_dates, by = "id") |>
  left_join(fresh_date, by = "id") |>
   filter(!is.na(date_fresh)) |>   #filter out na for fresh date
  mutate()

final_calfhealth

# Add new parameters to the tibble

final_calfhealth <- final_calfhealth |>
  mutate(
    event = !is.na(date_sold),
    end_date = if_else(event, date_sold, Sys.Date()),
    time_days = as.numeric(end_date - date_fresh)
  )

final_calfhealth |>
  group_by(disease_group) |>
  summarize(
    denominator = n(),
    left_herd = sum(event),
    percent_left = left_herd / denominator *100,
    .groups = "drop")

calf_summary <- final_calfhealth |>
  group_by(disease_group) |>
  summarize(
    denominator = n(),
    left_herd = sum(event),
    percent_left = left_herd / denominator * 100,
    .groups = "drop"
  )

ggplot(calf_summary, aes(x = disease_group, y = percent_left)) +
  geom_col(fill = "steelblue") +
  labs(
    x = "Disease Group",
    y = "Percent Left Herd (%)",
    title = "Percentage of Animals Leaving the Herd by Disease Group"
  ) +
  ylim(0, 100) +
  theme_minimal()

survival_summary <- final_calfhealth |>
  group_by(disease_group) |>
  summarise(
    animals_at_risk = n(),
    animals_left_herd = sum(event),
    animals_censored = sum(!event),
    percent_left_herd = animals_left_herd / animals_at_risk * 100,
    percent_remaining = animals_censored / animals_at_risk * 100,
    median_time_days = median(time_days, na.rm = TRUE),
    .groups = "drop"
  )

survival_summary

final_calfhealth <- final_calfhealth |>
    mutate(disease_group = factor(
      disease_group,
      levels = c("None", "Scours", "Pneu", "Both"),
      labels = c("None", "Scours", "Pneumonia", "Both")
    ))

fit <- survfit(
  Surv(time_days, event) ~ disease_group,
  data = final_calfhealth
)

ggsurvplot(
  fit,
  risk.table = TRUE,
  xlim = c(0, 1000),
  pval = TRUE,
  conf.int = FALSE,
  censor = FALSE
)

```






## Part 2 - Extension

Think about your disease or event of interest. Standard denominators like the one above are often not good enough.

Use your new data wrangling skills to create the perfect denominator for your question.


NOTE: Denominators can be challenging. If you get stuck:

-   First articulate what you want to do, step by step, in English

-   Then add a blank code chunk for each step

-   Begin filling in code

-   Ask the group for help if you get completely stuck.
