---
title: "Milestone 5 - Dates and Strings"
format: html
date: "`r Sys.Date()`"
embed-resources: true
---

```{r}
library(tidyverse)
library(quarto)
library(arrow)
library(waldo)

solution<-read_parquet('data/intermediate_files/events_formatted.parquet')

# read in files-----------------

list_files <- list.files(here::here("data/event_files")) # folder name where event files are located

events_original <- NULL

# i=1
for (i in seq_along(list_files)) {
  df <- read_csv(
    here::here(
      "data/event_files",
      list_files[i]
    ),
    # reads in all data as character string
    col_types = cols(.default = "c")
  )
  
   events_original <- bind_rows(events_original, df)
}

```

## Dates

The original dairy comp data is read into your environment and named "events_original".
For consistency all columns are read in as characters.

Use your new skills formating dates to create a new column named date_event that is a real date from the original column named Date.

Then create new variables that provide floor dates by month, season, and year. Name these floordate_month, floordate_year, floordate_season.

```{r initial denominator generation, message==FALSE}
events_formatted <- events_original |>
  mutate(
    date_event = mdy(Date),
    floordate_month = floor_date(date_event, unit="month"),
    floordate_year = floor_date(date_event, unit="year"),
    floordate_season = floor_date(date_event, unit="season")
  )

events_formatted |>
  select(Date, date_event, floordate_month, floordate_year, floordate_season) |>
  glimpse()
```

Do your new dates match the dates in the solution?


# Strings

Find all the columns in the solution that contain the word "remark".  

These are all based on the original Remark column in dairy comp. 

Can you recreate them?  

Hint: don't spend too much time trying to get exactly the same values. 
Just spend time thinking about how you could use them to clean up remarks in a way that would help you answer a question.


```{r, echo = TRUE}
names(solution)|>
  str_subset(regex("remark", ignore_case = TRUE))

```

# Extension - Your Turn

Start with events_original.  

Use your wrangling skils to fix date formats, and simplify free text to answer a question on this dairy.

You can choose any question you want. 

Here are some examples:

-  Which drug is most commonly used for mastitis treatment and has this changed over time?
-  What percent of breedings use sexed semen?  (this requires a large wrangle with an external key)
-  What specific lameness diagnosis are there and how frequently do they occur? 

```{r, echo = TRUE}
#Clean event dates & remarks
events_clean <- events_original |>
  mutate(
    date_event = mdy(Date),
    floordate_month = floor_date(date_event, unit="month"),
    floordate_year = floor_date(date_event, unit="year"),
    floordate_season = floor_date(date_event, unit="season"),
event_clean = Event |>
  str_to_upper()|>
  str_squish(),
protocols_clean = Protocols|>
  str_to_upper()|>
  str_squish()
  )

#pull out mastitis events
mastitis_events <-events_clean|>
  filter(str_detect(event_clean, "MAST|MASTITIS"))

#see list of potential mastitis treatments (actually under Protocols, Remarks has teat/quarter info)
mastitis_events |>
  count(protocols_clean, sort = TRUE)

#clean up protocols -  a couple random ones that I don't know what they are, so they're ambiguously renamed
mastitis_treatments <- mastitis_events |>
  mutate(
    mastitis_drug = case_when(
      is.na(protocols_clean) | protocols_clean == "" | protocols_clean == "NA" ~ "No protocol",
      str_detect(protocols_clean, "AHV PILL") ~ "AHV Pill",
      str_detect(protocols_clean, "CMT NEGATIVE") ~ "CMT Negative",
      str_detect(protocols_clean, "NO TREATMENT") ~ "No Treatment",
      str_detect(protocols_clean, "SPECTRA 3DAY") ~ "Spectramast LC x 3 days",
      str_detect(protocols_clean, "SPECTRA-5DY") ~ "Spectramast LC x 5 days",
      str_detect(protocols_clean, "SYSTEMIC MASTIT") ~ "Systemic antibiotics",
      str_detect(protocols_clean, "TR MASTITIS 1") ~ "Mastitis Treatment 1",
      str_detect(protocols_clean, "TODAY 3 DAY") ~ "Today x 3 days",
      str_detect(protocols_clean, "TODAY 5 DAY") ~ "Today x 5 days",
      str_detect(protocols_clean, "AMX3") ~ "Amoxi-Mast x 3 days",
      TRUE ~ "Other"
    )
  )

#this will show me all my mastitis treatments & count them
mastitis_treatments|>
  count(mastitis_drug, sort=TRUE)

#This just ensures I've captured everything in my prescribed categories - they're all in tehre! Nothing in teh Other category
mastitis_treatments |>
  filter(mastitis_drug == "Other") |>
  count(protocols_clean, sort = TRUE)

#let's plot it - I like a bar graph, but not stacked (position="dodge")
mastitis_treatments|>
  count(floordate_year, mastitis_drug)|>
  ggplot(aes(x=floordate_year, y=n, fill=mastitis_drug))+
  geom_col(position="dodge")+
  labs(
    title="Mastitis treatment protocols over time",
    x="Year",
    y="Number of mastitis events",
    fill="Treatment protocol"
  )

#let's plot it by facets - harder to interpret across treatments as separate graphs
mastitis_treatments|>
  count(floordate_year, mastitis_drug)|>
  ggplot(aes(x=floordate_year, y=n))+
  geom_col()+
  facet_wrap(vars(mastitis_drug),scales="free_y")+
  labs(
    title="Mastitis treatment protocols over time",
    x="Year",
    y="Number of mastitis events"
  )

#let's take into account how the dairy changed by switching to proportions
mastitis_treatments |>
  count(floordate_year, mastitis_drug) |> #compare year & drug
  group_by(floordate_year) |> #show points at each year
  mutate(
    total_mastitis_events = sum(n), 
    prop = n / total_mastitis_events #proportion
  ) |>
  ungroup() |>
  ggplot(aes(x = floordate_year, y = prop, color = mastitis_drug)) + #each drug gets its own color
  geom_point() +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Proportion of mastitis treatment protocols over time",
    x = "Year",
    y = "Percent of mastitis events",
    color = "Treatment protocol"
  )

```
